Linkage disequilibrium in conserved domains reveals opposing selective interactions in genes

### 2021-03-02 Manuscript.Rmd
add supp figures and msl dfe and mnm analysis

---
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: no
    number_sections: false
    keep_tex: true
    includes:
      in_header: "preamble.tex"
    citation_package: natbib
    extra_dependencies: ["float"]
bibliography: "references.bib"
biblio-style: "genetics"
title: Can we distinguish modes of selective interactions using linkage disequilibrium? ((or)) Mutation correlations in conserved domains reveal a heterogeneous landscape of selective interactions within genes ((or...?))
author:
header-includes:
- |
  ```{=latex}
  \usepackage{authblk}
  \author{Aaron P. Ragsdale \\ aaronpeaceragsdale@gmail.com}
  \affil{Department of Human Genetics, McGill University, Montreal, Canada \\
         National Laboratory of Genomics for Biodiversity, Irapuato, Mexico \\
         }
  ```
date: \today
---

@@

Most new mutations that affect fitness are deleterious and tend to be
eliminated from a population. The amount of time that a deleterious mutation
segregates in a population depends on the strength of selection against genomes
that carry it, with very damaging mutations kept at low frequencies in the
population and purged relatively rapidly. In the time between mutation and
fixation or loss, selected variants, both beneficial and damaging, can
dramatically impact patterns of variation in nearby linked regions. This
distortion away from neutral expectations has been widely documented in
sequencing data, but questions remain about the primary mode of interactions
between multiple linked variants and their joint effects on genome-wide
patterns of diversity.

@@

Non-additive effects, including dominance (i.e., interactions *within* loci)
and epistasis (interactions *between* loci), further complicate our
evolutionary models. A large fraction of nonsynonymous coding mutations are
thought to be at least partially recessive [@Agrawal2011-w; b@Huber2018-cp],
with average levels of dominance correlating with strength of selection
[@Kacser1981-nc], and dominance plays an important role in shaping expected
equilibrium allele frequencies and mutation load of strongly damaging disease
mutations [@Clark1998-kq]. On the other hand, epistasis differentially impacts
the deleterious load for asexually and sexually reproducing organisms
[@Kimura1966-gc; @Kondrashov1995-va], has been invoked as an explanation for
the evolutionary advantage of sex [@Kondrashov1982-sf; @Charlesworth1990-kw;
@Barton1998-lu], and can drive incompatibilities that lead to postzygotic
isolation during the process of speciation [@Turelli2000-kz]. Within
populations, epistasis is known to cause signed LD to deviate dramatically from
zero [@Charlesworth1990-kw; @Kondrashov1995-va]. However, despite dominance
being appreciated as important factors shaping linked variation
[@Turelli2000-kz; @Zhao2016-bb] and the evolutionary importance of epistatic
interactions, we currently lack models for predicting correlation patterns
between linked mutations under general selection models.

@@

A number of factors impede our interpretation of patterns of signed LD between
coding variants. First, for strongly deleterious or loss-of-function variants,
their rarity and low frequency means that statistical measurement of LD and
other diversity measures are very noisy. Second, comparisons are based on
theory that are confined to limiting and simplistic assumptions, including
steady-state demography, simple selection and interaction models, or unlinked
loci. To generate predictions under more complex models, we rely on expensive
forward simulations. These can be great for building intuition or testing
inference methods, but do not efficiently provide expectations for quantities
of interest across the range of relevant parameters. Analytical and numerical
methods for expected haplotype frequencies and LD under general selective
interaction models are thus crucial for interpreting patterns of variation
observed in data.

@@

In the Methods, I expand on the moment system developed in @Ragsdale2019-nt to
compute the expected sampling distribution $\Psi_n$ of two-locus haplotypes
under a general model of selective interactions. This sampling distribution, or
two-locus haplotype frequency spectrum, stores the expected density or observed
counts of pairs of loci with each possible haplotype frequency configuration in
a sample of size $n$. In many cases it is simpler to visualize summaries (such
as $\E[D]$) or slices of the distribution (Figure \@ref(fig:fig1)) instead of
the full three-dimensional distribution. Here, I focus on low-order LD
statistics, such as $\E[D]$ and $\E[D^2]$, and their decay with recombination
distance as these are statistics that are commonly used to test for
interactions between loci. Instead of unnormalized $\E[D^2]$ and $\E[D]$, I
consider expectations for $\sigma_d^2 = \E[D^2]/\E[p(1-p)q(1-q)]$ and
$\sigma_d^1 = \E[D]/\E[p(1-p)q(1-q)$. These statistics have two benefits:
first, the mutation rates cancel so that expectations are robust to assumptions
about the per-base mutation rate, and second, we have analytic expectations for
these quantities under neutrality at steady state for comparison
[@Ohta1971-yd].

I first consider the case of additive selection and epistasis, followed by
dominance acting within loci but no epistatic effects, and then describe a
general diploid selection model and consider simple simple gene-based dominance
effects. In what follows I focus on parameters where the strengths of selection
and dominance at each locus are equal, but note that the methods and
implementation allow for arbitrary and unequal selection and dominance at the
two loci.

@@

With epistasis, mean signed LD is large for both weakly and strongly selected
variants, with sign depending on the direction of epistatic interactions
(Figure \@ref(fig:fig2)). Synergistic epistasis results in negative LD while
antagonistic epistasis results positive LD, and large positive LD can occur
even when the epistatic effect is relatively weak. Epistasis-induced LD can
extend over long distances, especially for strongly deleterious mutations. For
example, in Figure \@ref(fig:fig1)D with $\gamma=-10$, which would be
considered moderately strong selection, large mean LD extends well beyond 100
kilobases, with more strongly deleterious interacting mutations expected to
show large signed LD over much larger recombination distances.

### Simple dominance within a locus

The effects of epistasis between variants on LD has been widely documented.
However, the role that non-additive selection within a locus affects
interference and LD between loci has received less attention. @Garcia2020-zn
used large-scale forward simulation to explore the effect of dominance on
patterns of LD, showing that LD depends on both the magnitude of the fitness
and dominance effects in a nonlinear way, but note that expensive large-scale
simulations are needed to explore these effects as no closed form approach to
computing those expectations exist. The method presented here provides such a
closed form approach, so that expectations of signed LD and other two-locus
statistics can be computed more efficiently and accurately.

The combined effect of the strength of selection and dominance on interference
is indeed nontrivial, and some parameter regimes can cause strong negative LD
between pairs of negatively selected variants. For strong selection, pairs of
recessive mutations ($h<1/2$) can have large negative signed LD (Figure
@\ref(fig:fig3)D), although it decays rapidly with increasing distance between
loci and is roughly zero for $\rho > 1$. For weakly selected mutations with
$\gamma=-1$, we see that there is no monotonic effect of the level of dominance
on negative LD, with both recessive and dominant pairs of mutations having more
negative LD than the additive case.

@@

The moment system for $\Psi_n$ readily incorporates variable population size or
mutation and recombination rates or selection parameters that change over time.
Here I focus on non-equilibrium population size history and consider scenarios
relevant to human demographic history, including bottlenecks and expansions. I
explore two simple models (Figure \@ref(fig:fig4)A), one with an instantaneous
expansion and another with a bottleneck followed by recovery. I also consider
two demographic histories inferred using genome-wide genealogy reconstruction
[@Speidel2019-nj] applied to the @1000_Genomes_Project_Consortium2015-zq
dataset, and focus on size histories for the Yoruba from Ibidan, Nigeria (YRI)
and Utahns of North and West European ancestry (CEU) (Figure \@ref(fig:fig4)B).

@@

I also measure LD between pairs of mutations that both fall outside of
annotated domains. For such mutations, I consider averages across all pairs
outside of domains, as well as subsets of pairs matched to the same distances
for pairs of mutation within and between domains (Figures \@ref(fig:fig6)B-D and
\@ref(fig:domainspops)). Averaging over all pairs of mutations within the same
gene but outside domains, both synonymous and missense mutations had LD
slightly above zero, with synonymous mutations observed to have slightly higher
signed LD than missense mutations. This pattern is primarily driven by tightly
linked variants. For pairs of mutations at distances matched to those within
the domains, synonymous mutations have much larger positive LD than
missense mutations, a signal that vanishes for pairs matched to distances
between domains. For larger distances, both synonymous and missense mutations
have LD approximately zero.

@@

Previous theoretical and simulation studies have shown that interference and
interactions between selected mutations reduces the efficacy of selection at
linked loci [@Hill1968-vu; @Birky1988-jm; @Barton1995-mj; @McVean2000-ox],
impacting substitution rates, the deleterious mutation load, and dynamics of
segregating mutations. Interference between moderately deleterious mutations
and synergistic epistasis are both expected to cause negative LD between
selected mutations, which can be readily tested using population genetic data
[@Sohail2017-zq; @Sandler2020-vk; @Garcia2020-zn]. Here, I first used a closed
numerical approach to generate expectations for signed LD under a wide range of
selective scenarios, and then compared patterns of LD in human populations
between classes of coding mutations within genes using unbiased estimators for
LD from unphased genotypes.

@@

From gene-wide averages, LD between pairs of missense mutations does not appear
to be different from pairs of synonymous variants, but LD between
loss-of-function variants is substantially more negative. Hill-Robertson
effects under additive selection are expected to be strongest for slightly to
moderately deleterious variants (with $s\sim 1/N_e$), as strongly deleterious
mutations are not expected to interfere with one another [@McVean2000-ox].
However, loss-of-function mutations do not typically fall within this regime,
and inference of the distribution of fitness effects (DFE) for new
loss-of-function variants shows that a large majority are strongly deleterious
(Supporting Information). Instead, negative synergistic epistasis between
strongly deleterious mutations does produce large negative deviations of mean
LD. Strongly deleterious *recessive* mutations can also produce this pattern,
making it difficult to distinguish these effects with certainty. This
difficulty is compounded by the large measurement noise for $\E[D]$, especially
for loss-of-function variants for which only a few hundred pairs are found
within genes in each population. If loss-of-function variants are in fact
recessive, even partially, this can have a large impact on the inferences of
the DFE (Supporting Information).

## Non-uniform interactive effects of selected mutations within genes

@@

## Clustered mutations can account for positive LD among synonymous mutations

@@

Nonrandom mutational processes provide an alternative explanation for positive
LD between neutral variants separated by short distances, in which clusters of
mutations occur simultaneously in the same mutational event. Such
multinucleotide mutations have been shown to affect patterns of LD over short
distances, on the order of 10s to 100s of base pairs [@Harris2014-zg], and
clustered mutational events may be common in humans [@Besenbacher2016-ac].
Thus, positive LD may be the baseline expectation for tightly linked variants
primarily due to clustered mutations, and only subsequent selective processes and
interactions will cause LD to deviate from that expectation, as seen for
missense mutations across different sub-elements of genes. Similarly, it may be
more appropriate to compare the negative LD observed between linked
loss-of-function variants to that positive baseline expectation instead of
zero, implying that they would be more recessive or have stronger synergistic
epistasis than if the neutral expectation were zero.

@@

In exploring or inferring multi-locus selective models, the number of plausible
selection scenarios becomes quite large as we relax the strict assumptions of
additive and multiplicative fitness effects. The inclusion of dominance
effects, epistasis, or other interactions leads to a rapid increase in the
number of parameters to consider. This makes performing forward simulations
that span the range of all such selective interaction scenarios burdensome.

@@

Finally, exploring additional summaries of the two-locus sampling distribution
could provide additional power to distinguish between interaction models. The
method developed here allows us to compute the full sampling distribution for a
given sample size, and expectations for a large family of informative two-locus
statistics can be computed directly from this distribution. These expectations
can then be compared directly to empirical observations, which can be taken
from either phased or unphased [@Ragsdale2020-nz]. Additional work performing
such an analysis is therefore warranted, which should provide a path forward
for distinguishing between modes of selective interactions.

@@

To include selection, however, there have been relatively few advances beyond
the Monte Carlo simulation approach taken by @Hill1966-gv, albeit we now have
access to somewhat more powerful computational resources and more sophisticated
software for performing flexible forward simulation (e.g., @Thornton2014-pn,
@Haller2019-vm). Analytic results for two-locus distributions under selection
are notoriously difficult, with a few notable flashes of progress.
@McVean2007-xd considered the effect of a recent sweep on patterns of LD for
loci nearby or spanning the locus under selection, and in an impressive recent
paper, @Good2020-wj presented analytic solutions for patterns of LD between
rare mutations under additive selection with epistasis. Nonetheless, such
approaches are confined to steady-state demography and constrained selection
models.

Numerical methods straddle the two approaches of expensive and noisy discrete
simulation and analytic solutions, providing a more efficient and practical
method to compute expectations of two-locus diversity measures under a wider
range of parameters and demographic scenarios. @Ragsdale2017-gg used a finite
differencing approach to numerically solve the two-locus diffusion equation
with additive selection at either locus, although that paper focused on the
applicability of two-locus statistics to demographic inference. More recently,
@Ragsdale2019-nt extended the @Hill1968-vu system for $\E[D^2]$ to compute
arbitrary moments of the distribution of $D$ for any number of populations
connected by migration and admixture. This paper also showed that such a
moments-construction can be used to solve for the two-locus sampling
distribution within a single population, though requires a moment-closure
approximation for nonzero recombination and selection. Below, I
extend this same approach to model arbitrary diploid selection, which
encompasses dominance, epistasis, and other forms of selective interactions
between two loci.

@@

The two-locus sampling distribution is the direct analog to the single-locus
site frequency spectrum (SFS) of a given sample size. Instead of describing the
density or number of mutations with a given allele count out of $n$ samples,
the two-locus distribution $\Psi_n$ stores the density or number of pairs of
loci with observed haplotype counts, so that $\Psi_n(i, j, k)$ is the number of
pairs where with $i$ copies of the $AB$ haplotype, $j$ of type $Ab$, $k$ of
type $aB$, and $n-i-j-k$ of type $ab$. The size of $\Psi_n$ grows rapidly, with
$O(n^3)$ entries, which practically limits computational approaches to moderate
sample sizes and a single population.

@@

This requirement of extra lineages for nonzero recombination and selection
means that the system in \@ref(eq:system) is not closed.
This means we need a moment-closure approximation to solve for $\Psi_n$. As in
@Ragsdale2019-nt, a jackknife approximation is used to estimate $\Psi_{n+l}$,
for $l=1$ or $2$, from $\Psi_n$ [@Jouganous2017-pq], so that
$\hat{\Psi}_{n+l}(i, j, k) = \mathcal{J}_{n, l}\Psi_n$, although other closure approximations are possible.
This gives us a closed approximated system,
\begin{equation}
(\#eq:closed-system)
\dot{\hat\Psi}_n(i, j, k; t) = 
\mathcal{D}_{\nu(t)}\hat\Psi_n(t)
+ \mathcal{R}_{\rho}\mathcal{J}_{n, 1}\hat\Psi_{n}(t)
+ \mathcal{U}_{\theta}\hat\Psi_n(t)
+ \mathcal{S}_{\gamma, h}\mathcal{J}_{n, 2}\hat\Psi_{n}(t).
\end{equation}

The jackknife approximation, which approximates an entry $\Psi_{n+l}(i,j,k)$
using nearby entries in $\Psi_n$, is more accurate for larger sample sizes,
creating a tension between efficiency and accuracy: larger sample sizes result
in more accurate solutions, as error in the jackknife is diminished, but
computational complexity also grows rapidly in the number of entries of
$\Psi_n$, which is $O(n^3)$. Accuracy and runtime are shown in Figure
\@ref(fig:jackknife). In the results presented in this paper, sample sizes
between $n=30$ and $n=80$ were used.

@@

To model selection, we consider a model where we draw lineages uniformly from
the previous generation, but keep lineages with probability proportional to
their fitness. In the absence of dominance, the selection model reduces to a
haploid model, with acceptance and rejection probabilities depending on the
fitnesses of each haploid copy, where haplotype $Ab$ has fitness $1 + s_A$,
$aB$ has fitness $1 + s_B$, and $AB$ has fitness $s_{AB}$. We assume the doubly
ancestral haplotype $ab$ has fitness $1$, so fitnesses are relative to that of
$ab$ haplotypes. The standard multiplicative fitness function assumes that
$s_{AB} \approx s_A + s_B$, and a general epistasis function can be written as
$s_{AB} = (s_A + s_B) (1 + \epsilon)$, so that $\epsilon>0$ implies synergistic
epistasis, while $\epsilon < 0$ implies antagonistic epistasis.

@@

# Acknowledgements
- Simon Gravel
- Kevin Thornton
- Alex Diaz-Papkovich
- Mashaal Sohail

>>>>

---
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: no
    number_sections: false
    keep_tex: true
    includes:
      in_header: "preamble.tex"
    citation_package: natbib
    extra_dependencies: ["float"]
bibliography: "references.bib"
biblio-style: "genetics"
title: Linkage disequilibrium in conserved domains reveals opposing selective interactions in genes
author:
header-includes:
- |
  ```{=latex}
  \usepackage{authblk}
  \author{Aaron P. Ragsdale \\ aaronpeaceragsdale@gmail.com}
  \affil{Department of Human Genetics, McGill University, Montreal, Canada \\
         National Laboratory of Genomics for Biodiversity, Irapuato, Mexico \\
         }
  ```
date: \today
---

@@

Most new mutations that affect fitness are deleterious and tend to be
eliminated from a population. The amount of time that a deleterious mutation
segregates depends on the strength of selection against genomes that carry it,
with very damaging mutations kept at low frequencies and purged relatively
rapidly. In the time between mutation and fixation or loss, selected variants,
both beneficial and damaging, can dramatically impact patterns of variation in
nearby linked regions. This distortion away from neutral expectations has been
widely documented in sequencing data, but questions remain about the primary
mode of interactions between multiple linked variants and their joint effects
on genome-wide patterns of diversity.

@@

Non-additive effects, including dominance (i.e., interactions *within* loci)
and epistasis (interactions *between* loci), further complicate our
evolutionary models. A large fraction of nonsynonymous coding mutations are
thought to be at least partially recessive [@Agrawal2011-wb; @Huber2018-cp],
with average levels of dominance correlating with strength of selection
[@Kacser1981-nc], and dominance plays an important role in shaping expected
equilibrium allele frequencies and mutation load of strongly damaging disease
mutations [@Clark1998-kq]. On the other hand, epistasis differentially impacts
the deleterious load for asexually and sexually reproducing organisms
[@Kimura1966-gc; @Kondrashov1995-va], has been invoked as an explanation for
the evolutionary advantage of sex [@Kondrashov1982-sf; @Charlesworth1990-kw;
@Barton1998-lu], and can drive incompatibilities that lead to postzygotic
isolation during the process of speciation [@Turelli2000-kz]. Within
populations, epistasis is known to cause signed LD to deviate dramatically from
zero [@Charlesworth1990-kw; @Kondrashov1995-va]. However, despite dominance
being appreciated as important factors shaping linked variation
[@Turelli2000-kz; @Zhao2016-bb] and the evolutionary importance of epistatic
interactions, we currently lack models for predicting correlation patterns
between linked mutations under general selection models.

@@

A number of factors impede our interpretation of patterns of signed LD between
coding variants. First, for strongly deleterious or loss-of-function variants,
their rarity and low frequency means that statistical measurement of LD and
other diversity measures are very noisy. Second, comparisons are based on
theory that are confined to limiting and simplistic assumptions, including
steady-state demography, simple selection and interaction models, or unlinked
loci. To generate predictions under more complex models, we rely on expensive
forward simulations. Such simulations can help build intuition and test
inference methods, but do not efficiently provide expectations for quantities
of interest across the range of relevant parameters. Analytical and numerical
methods for expected haplotype frequencies and LD under general selective
interaction models are thus crucial for interpreting patterns of variation
observed in data.

@@

In the Methods, I expand on the moment system developed in @Ragsdale2019-nt to
compute the expected sampling distribution $\Psi_n$ of two-locus haplotypes
under a general model of selective interactions. This sampling distribution, or
two-locus haplotype frequency spectrum, stores the expected density or observed
counts of pairs of loci with each possible haplotype frequency configuration in
a sample of size $n$. In many cases it is simpler to visualize summaries (such
as $\E[D]$) or slices of the distribution (Figure \@ref(fig:fig1)) instead of
the full three-dimensional distribution. Here, I focus on low-order LD
statistics, such as $\E[D]$ and $\E[D^2]$, and their decay with recombination
distance as these are statistics that are commonly used to test for
interactions between loci. Instead of unnormalized $\E[D^2]$ and $\E[D]$, I
consider expectations for $\sigma_d^2 = \E[D^2]/\E[p(1-p)q(1-q)]$ and
$\sigma_d^1 = \E[D]/\E[p(1-p)q(1-q)]$. These statistics have two benefits:
first, the mutation rate cancels so that expectations are robust to assumptions
about the per-base mutation rate, and second, analytic expectations for these
quantities are available under neutrality at constant population size for
comparison [@Ohta1971-yd].

I first consider the case of additive selection and epistasis, followed by
dominance acting within loci but no epistatic effects, and then describe a
general diploid selection model and consider simple gene-based dominance
effects. In what follows I focus on parameters where the strengths of selection
and dominance at each locus are equal, but note that the methods and
implementation allow for arbitrary and unequal selection and dominance at the
two loci.

@@

With epistasis, mean signed LD is large for both weakly and strongly selected
variants, with sign depending on the direction of epistatic interactions
(Figure \@ref(fig:fig2)). Synergistic epistasis results in negative LD while
antagonistic epistasis results positive LD, and large positive LD can occur
even when the epistatic effect is relatively weak. Epistasis-induced LD can
extend over long distances, especially for strongly deleterious mutations. For
example, in Figure \@ref(fig:fig1)D with $\gamma=-10$, which would be
considered moderate, large mean LD extends well beyond 100 kilobases, with more
strongly deleterious interacting mutations expected to show large signed LD
over much larger recombination distances.

### Dominance and interference

The effects of epistasis between variants on LD has been widely documented.
However, the effect of non-additive selection within a locus on interference
has received less attention. @Garcia2020-zn used large-scale forward simulation
to explore the effect of dominance on patterns of LD, showing that LD depends
on both the magnitude of the fitness and dominance effects in a nonlinear way,
but note that expensive large-scale simulations are needed to explore these
effects as no closed form approach to computing those expectations exist. The
method presented here provides such a closed form approach, so that
expectations of signed LD and other two-locus statistics can be computed more
efficiently and accurately.

The combined effect of the strength of selection and dominance on interference
is indeed nontrivial, and some parameter regimes can cause strong negative LD
between pairs of negatively selected variants. For strong selection, pairs of
recessive mutations ($h<1/2$) can have large negative signed LD (Figure
\@ref(fig:fig3)D), although it decays rapidly with increasing distance between
loci and is roughly zero for $\rho > 1$. For weakly selected mutations with
$\gamma=-1$, we see that there is no monotonic effect of the level of dominance
on negative LD, with both recessive and dominant pairs of mutations having more
negative LD than the additive case.

@@

The moment system for $\Psi_n$ readily incorporates variable population size as
well as mutation and recombination rates and selection parameters that change
over time. Here I focus on non-equilibrium population size history and consider
scenarios relevant to human demographic history, including bottlenecks and
expansions. I explore two simple models (Figure \@ref(fig:fig4)A), one with an
instantaneous expansion and another with a bottleneck followed by recovery. I
also consider two demographic histories inferred using genome-wide genealogy
reconstruction [@Speidel2019-nj] applied to the
@1000_Genomes_Project_Consortium2015-zq dataset, and focus on size histories
for the Yoruba from Ibidan, Nigeria (YRI) and Utahns of North and West European
ancestry (CEU) (Figure \@ref(fig:fig4)B).

@@

I also measure LD between pairs of mutations that both fall outside of
annotated domains. For such mutations, I consider averages across all pairs
outside of domains, as well as subsets of pairs matched to the same distances
for pairs of mutation within and between domains (Figures \@ref(fig:fig6)B-D and
\@ref(fig:domainspops)). Averaging over all pairs of mutations within the same
gene but outside domains, both synonymous and missense mutations have LD
slightly above zero, with synonymous mutations observed to have slightly higher
signed LD than missense mutations. This pattern is primarily driven by tightly
linked variants. For pairs of mutations at distances matched to those within
the domains, synonymous mutations have much larger positive LD than
missense mutations, a signal that vanishes for pairs matched to distances
between domains. For larger distances, both synonymous and missense mutations
have LD approximately zero.

@@

Previous theoretical and simulation studies have shown that interference and
interactions between selected mutations reduce the efficacy of selection at
linked loci, impacting substitution rates, the deleterious mutation load, and
dynamics of segregating mutations [@Hill1968-vu; @Birky1988-jm; @Barton1995-mj;
@McVean2000-ox]. Interference between moderately deleterious mutations and
synergistic epistasis are both expected to cause negative LD between selected
mutations, which can be readily tested using population genetic data
[@Sohail2017-zq; @Sandler2020-vk; @Garcia2020-zn]. Here, I first used a closed
numerical approach to generate expectations for signed LD under a wide range of
selective scenarios, and then compared patterns of LD in human populations
between classes of coding mutations within genes using unbiased estimators for
LD from unphased genotypes.

@@

From gene-wide averages, LD between pairs of missense mutations does not appear
to be different from pairs of synonymous variants, but LD between
loss-of-function variants is substantially more negative. Hill-Robertson
effects under additive selection are expected to be strongest for slightly to
moderately deleterious variants (with $s\sim 1/N_e$), as strongly deleterious
mutations are not expected to interfere with one another [@McVean2000-ox].
However, loss-of-function mutations do not typically fall within this regime,
and inference of the distribution of fitness effects (DFE) for new
loss-of-function variants shows that a large majority are strongly deleterious
(Supporting Information). Instead, negative synergistic epistasis between
strongly deleterious mutations does produce large negative deviations of mean
LD. Strongly deleterious *recessive* mutations can also produce this pattern,
making it difficult to distinguish these effects with certainty. This
difficulty is compounded by the large measurement noise for $\E[D]$, especially
for loss-of-function variants for which only a few hundred pairs are found
within genes in each population. If loss-of-function variants are in fact
recessive, even partially, this can impact inferences of
the DFE (Supporting Information).

## Non-uniform interactions between selected mutations within genes

@@

## Clustered mutations account for positive LD among synonymous mutations

@@

Nonrandom mutational processes provide an alternative explanation for positive
LD between neutral variants separated by short distances, in which clusters of
mutations occur simultaneously in the same mutational event. Such
multinucleotide mutations have been shown to affect patterns of LD over short
distances, on the order of 10s to 100s of base pairs [@Harris2014-zg], and
clustered mutational events may be common in humans [@Besenbacher2016-ac].
Indeed, a simple exponential model in which the fraction of pairs of mutations
arising arising from a multinucleotide event decays with distance fits the
observed decay of $\sigma_d^1$ between synonymous mutations (Supporting
Information). We may therefore treat positive LD as the baseline expectation
for tightly linked variants primarily due to clustered mutations, so that
subsequent selective processes and interactions cause LD to deviate from that
expectation, as seen for missense mutations across different sub-elements of
genes. It may also be more appropriate to compare the negative LD observed
between linked loss-of-function variants to that positive baseline expectation
instead of zero, so that they are more recessive or have stronger synergistic
epistasis than if the neutral expectation were zero.

@@

In exploring or inferring multi-locus selective models, the number of plausible
selection scenarios becomes quite large as we relax the strict assumptions of
additivity and multiplicative fitness effects. The inclusion of dominance
effects, epistasis, or other interactions leads to a rapid increase in the
number of parameters to consider. This makes performing forward simulations
that span the range of all such selective interaction scenarios burdensome.

@@

Finally, exploring additional summaries of the two-locus sampling distribution
could provide additional power to distinguish between interaction models. The
method developed here allows us to compute the full sampling distribution for a
given sample size, and expectations for a large family of informative two-locus
statistics can be computed directly from this distribution. These expectations
can then be compared directly to empirical observations, which can be taken
from either phased or unphased data [@Ragsdale2020-nz]. Additional work
performing such an analysis is therefore warranted, which should provide a path
forward for distinguishing between modes of selective interactions.

@@

To include selection, however, there have been relatively few advances beyond
the Monte Carlo simulation approach taken by @Hill1966-gv, albeit we now have
access to somewhat more powerful computational resources and more sophisticated
software for performing flexible forward simulation (e.g., @Thornton2014-pn,
@Haller2019-vm). Analytic results for two-locus distributions under selection
are notoriously difficult, with a few notable flashes of progress. For example,
@McVean2007-xd considered the effect of a recent sweep on patterns of LD in
loci nearby or spanning the locus under selection, and in an impressive recent
paper, @Good2020-wj presented analytic solutions for patterns of LD between
rare mutations under additive selection with epistasis. Nonetheless, such
approaches are typically confined to steady-state demography and constrained
selection models.

Numerical methods inhabit the space in between expensive discrete simulations
and limited analytic solutions, providing a more efficient and practical method
to compute expectations of two-locus diversity measures under a wider range of
parameters and demographic scenarios. @Ragsdale2017-gg used a finite
differencing approach to numerically solve the two-locus diffusion equation
with additive selection at either locus, although that paper focused on the
applicability of two-locus statistics to demographic inference. More recently,
@Ragsdale2019-nt extended the @Hill1968-vu system for $\E[D^2]$ to compute
arbitrary moments of the distribution of $D$ for any number of populations
connected by migration and admixture. This paper also showed that such a
moments-construction can be used to solve for the two-locus sampling
distribution within a single population, though it requires a moment-closure
approximation for nonzero recombination and selection. Below, I extend this
approach to model arbitrary diploid selection, which encompasses dominance,
epistasis, and other forms of selective interactions between two loci.

@@

The two-locus sampling distribution is the direct analog to the single-locus
site frequency spectrum (SFS) of a given sample size. Instead of describing the
density or number of mutations with a given allele count out of $n$ samples,
the two-locus distribution $\Psi_n$ stores the density or number of pairs of
loci with observed haplotype counts, so that $\Psi_n(i, j, k)$ is the number of
pairs for which we observe $i$ copies of the $AB$ haplotype, $j$ of type $Ab$,
$k$ of type $aB$, and $n-i-j-k$ of type $ab$. The size of $\Psi_n$ grows
rapidly, with $O(n^3)$ entries, which practically limits computational
approaches to moderate sample sizes and a single population.

@@

This requirement of extra lineages for nonzero recombination and selection
means that the system in \@ref(eq:system) is not closed, so that we need a
moment-closure approximation to solve for $\Psi_n$. As in @Ragsdale2019-nt, a
jackknife approximation is used to estimate $\Psi_{n+l}$, for $l=1$ or $2$,
from $\Psi_n$ [@Jouganous2017-pq], so that $\hat{\Psi}_{n+l}(i, j, k) =
\mathcal{J}_{n, l}\Psi_n$, although other closure approximations are possible.
This emits a closed approximate system,
\begin{equation}
(\#eq:closed-system)
\dot{\hat\Psi}_n(i, j, k; t) = 
\mathcal{D}_{\nu(t)}\hat\Psi_n(t)
+ \mathcal{R}_{\rho}\mathcal{J}_{n, 1}\hat\Psi_{n}(t)
+ \mathcal{U}_{\theta}\hat\Psi_n(t)
+ \mathcal{S}_{\gamma, h}\mathcal{J}_{n, 2}\hat\Psi_{n}(t).
\end{equation}

The jackknife approximation, which approximates an entry $\Psi_{n+l}(i,j,k)$
using nearby entries in $\Psi_n$, is more accurate for larger sample sizes,
creating a tension between efficiency and accuracy: larger sample sizes result
in more accurate solutions, as error in the jackknife is diminished, but
computational complexity also grows rapidly in the number of entries of
$\Psi_n$, which is $O(n^3)$. Accuracy and runtime are shown in Figure
\@ref(fig:jackknife). In the results presented in this paper, sample sizes
between $n=30$ and $n=80$ are used.

@@

To include selection, we consider a model where we draw lineages uniformly from
the previous generation, but keep lineages with probability proportional to
their fitness. In the absence of dominance, selection reduces to a haploid
model, with acceptance and rejection probabilities depending on the fitnesses
of each haploid copy, where haplotype $Ab$ has fitness $1 + s_{A}$, $aB$ has
fitness $1 + s_{B}$, and $AB$ has fitness $1 + s_{AB}$. We assume the doubly
ancestral haplotype $ab$ has fitness $1$, so fitnesses are relative to that of
$ab$ haplotypes. The standard multiplicative fitness function assumes that
$s_{AB} \approx s_{A} + s_{B}$, and a general epistasis function can be written
as $s_{AB} = (s_{A} + s_{B}) (1 + \epsilon)$, so that $\epsilon > 0$ implies
synergistic epistasis, while $\epsilon < 0$ implies antagonistic epistasis.

@@

# Acknowledgements
- Simon Gravel
- Kevin Thornton
- Alex Diaz-Papkovich
- Mashaal Sohail
- Matthias Steinrücken
